Supplementary Information

### 2021-04-06 doc/manuscript/ForestGEO_treerings.Rmd
finish NP feedback

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
These findings are generally consistent with the global-scale analysis of @babst_twentieth_2019, which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that in most forests tree growth is likely to be reduced by warming temperatures.

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the primary climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
~~This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.~~ *(2nd sentence moved to methods)*

Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent (Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent [@germain_climate_2020].

@@

Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`). 
Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theorydriven_2008].
However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach **conceptually?** parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted with caution, and as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

The majority of negative growth trends observed here (Fig. `r year_fig`) are probably attributable to stand dynamics as cohorts and stands age, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs. `r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand. 
This can occur, for example, when a species undergoes a dramatic decline in recruitment within a stand (e.g., *Quercus* spp. at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger DBH classes as time proceeds. 
In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs. `r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
For more shade-tolerant species in stands with no known major disturbance within the past 150 years (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].

In a few instances, directional changes in growth are likely attributable to abiotic drivers.
In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_longterm_2008], as captured in our records (Fig. `r decadal_plot_ZOF`), and this influenced linear growth trends of various species differently within the GLS model.
In cases such as this, a non-linear response function to yeaar, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

>>>>

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
These findings are generally consistent with the global-scale analysis of @babst_twentieth_2019, which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that warmer temperatures are likely to reduce growth at most of the sites analyzed here, in large part by exacerbating observed water limitations (**REFS**).

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the primary climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged (**REFS (NEIL)**). 
Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
~~This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.~~ *(2nd sentence moved to methods)*

Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent (Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].

@@

Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`). 
Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theorydriven_2008].
However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach **conceptually?** parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

The majority of negative growth trends observed here (Fig. `r year_fig`) are probably attributable to stand dynamics as cohorts and stands develop oer time, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs. `r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand. 
This can occur, for example, when a species undergoes a dramatic decline in recruitment within a stand (e.g., *Quercus* spp. at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger DBH classes as time proceeds. 
In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs. `r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
For more shade-tolerant species in stands with no known major disturbance within the past 150 years (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].

In a few instances, directional changes in growth are likely attributable to abiotic drivers.
In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_longterm_2008], as captured in our records (Fig. `r decadal_plot_ZOF`), and this influenced linear growth trends of various species differently within the GLS model.
In cases such as this, a non-linear response function to year, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
